Agda Meets Accelerate

### 2012-08-30 talk/ifl2012.tex
draft

\begin{frame}
  \frametitle{Introduction}
  \begin{itemize}
  \item Embedded languages
  \item Dependent types
  \item Generative programming
  \item Data-parallel programming
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Overview}
  \begin{block}{Mission}
    Construct a generative, data parallel programming language embedded in Agda
  \end{block}
  \begin{block}<2->{Means}
    \dots\ by creating a --- more expressive --- Agda frontend for the
    Accelerate language.
  \end{block}
\end{frame}

\section{Accelerate}
\begin{frame}
  \frametitle{The Accelerate Language}
  \begin{itemize}
  \item Data-parallel array language
  \item Expressive embedding in Haskell
  \item Generates GPU code at run time
  \end{itemize}
\end{frame}

@@

\begin{frame}[fragile]
  \frametitle{Types in Accelerate}
  \begin{itemize}
  \item \texttt{Vector} and \texttt{Scalar} are suitable
    instantiations of a parameterized array type:
\begin{verbatim}
type Scalar e = Array DIM0 e
type Vector e = Array DIM1 e
\end{verbatim}
  \item Dimensions in turn are abbreviations for certain array shapes:
\begin{verbatim}
type DIM0 = Z
type DIM1 = DIM0 :. Int
-- <and so on>
\end{verbatim}
  \item and simple shapes are built from two basic constructors
\begin{verbatim}
data Z       = Z
data sh :. i = sh :. i
\end{verbatim}
  \end{itemize}
\end{frame}

@@

\begin{frame}[fragile]
  \frametitle{Indexed Types}
  \begin{itemize}
  \item Standard example: vector type indexed by its size
\begin{verbatim}
data Vec (A : Set) : Nat -> Set where
  []   : Vec A zero
  _::_ : {n : Nat} -> A -> Vec A n -> Vec A (suc n)
\end{verbatim}
  \item Enables the construction of statically safe access operations
    for vectors
\begin{verbatim}
get : forall {A : Set} {n : Nat} ->
      Vec A n -> (m : Nat) -> m < n -> A
get []        _       ()      -- impossible case
get (x :: xs) zero    p       = x
get (x :: xs) (suc m) (s<s p) = get xs m p
\end{verbatim}
  \end{itemize}
\end{frame}

@@

\begin{frame}[fragile]
  \frametitle{Example: Dot Product in Agda-Accelerate}
  \vspace{-\baselineskip}
\begin{verbatim}
dotp : forall {E : Element} {{p : IsNumeric E}} {n : Nat}
     -> PreVector n E -> PreVector n E -> Scalar E
dotp{E} xs ys = 
  let xs' = use xs
      ys' = use ys
  in
  fold _+_ ("0" ::: E) (zipWith _*_ xs' ys')
\end{verbatim}
  \begin{itemize}
  \item \texttt{PreVector n E} : a vector of $n$ elements of
    type \texttt{E}
  \item \texttt{Scalar E} : code that evaluates to a scalar of type \texttt{E}
  \item \texttt{IsNumeric E} restricts \texttt{E} to numeric types
  \end{itemize}
\end{frame}
\begin{frame}[fragile]
  \frametitle{The Array Type}
  \begin{itemize}
  \item The array type is indexed over a shape
  \item Shapes contain size information
\begin{verbatim}
data Shape : Set where
  Z     : Shape
  _:<_> : Shape -> Nat -> Shape
\end{verbatim}
  \item \texttt{Vector} and \texttt{Scalar} are, again,
    specializations of a shape-indexed array type
\begin{verbatim}
Vector n E    = Array    (Z :< n >) E
Scalar E      = Array    Z          E
\end{verbatim}
  \end{itemize}
\end{frame}
\begin{frame}[fragile]
  \frametitle{Bounds Checking: zipWith}
\begin{verbatim}
zipWith : forall {A B C} {sh} -> (Exp A -> Exp B -> Exp C)
        -> Array sh A -> Array sh B -> Array sh C
\end{verbatim}
  \begin{itemize}
  \item Ensures that only arrays of the exact same shape and size are
    zipped together
  \item More restrictive than Accelerate that only checks the
    dimension and then takes the minimum
  \item But that can also be encoded
  \end{itemize}
\end{frame}
\begin{frame}[fragile]
  \frametitle{Accelerate's zipWith Function in Agda}
\begin{verbatim}
zipWith' : forall {A B C} {sha shb} {p : EqShape sha shb}
        -> (Exp A -> Exp B -> Exp C)
        -> Array sha A -> Array shb B
        -> Array (shmin sha shb {p}) C
\end{verbatim}
where\scriptsize
\begin{verbatim}
data EqShape : Shape -> Shape -> Set where
  EqZ : EqShape Z Z
  EqS : {s1 s2 n1 n2} -> EqShape s1 s2 -> EqShape (s1 :< n1 >) (s2 :< n2 >)

@@

\begin{frame}[fragile]
  \frametitle{Bounds Checking: fold and fold1}

  \begin{itemize}
  \item \texttt{fold} and \texttt{fold1} take a binary operation and
    reduce one dimension of an array using the operation
  \item \texttt{fold} works for any array, but \texttt{fold1} requires
    a non-empty array argument
  \item The latter restriction is not expressed in Haskell-Accelerate
  \end{itemize}
  \begin{block}<2->{\vspace{-\baselineskip}}\small
\begin{verbatim}
fold  :: (Shape ix, Elt a) =>
         (Exp a -> Exp a -> Exp a) -> Exp a ->
         Acc (Array (ix :. Int) a) -> Acc (Array ix a)
fold1 :: (Shape ix, Elt a) =>
         (Exp a -> Exp a -> Exp a) ->
         Acc (Array (ix :. Int) a) -> Acc (Array ix a)
\end{verbatim}
  \end{block}
\end{frame}

@@

fold1 : forall {E}{sh}{n}
      -> (Exp E -> Exp E -> Exp E)
      -> Array (sh :< suc n >) E
      -> Array sh E
\end{verbatim}
\end{frame}
\begin{frame}[fragile]
  \frametitle{Another Issue with fold and fold1}
  \begin{itemize}
  \item \texttt{fold f e} and \texttt{fold1 f}  partition their array
    argument and apply \texttt{f} in a bracketing that is a-priori unknown
  \item For deterministic results, \texttt{f} must be associative and
    (in case of \texttt{fold}) \texttt{e} must be a unit of \texttt{f}
  \item<2-> Such restrictions can be expressed in Agda
  \item<3-> Might ask for a signature like this
\begin{verbatim}
fold : ... -> Monoid E
     -> Array (sh :< n >) E -> Array sh E
\end{verbatim}
  \end{itemize}
\end{frame}
\begin{frame}[fragile]
  \frametitle{A Monoidal Catch}
An encoding of monoids in Agda:\scriptsize
\begin{verbatim}
record Monoid (E : Set) : Set where
  field
    op : E -> E -> E
    un : E
    op_assoc : forall {x y z} -> op (op x y) z = op x (op y z)
    op_un_le : forall {x} -> op un x = x
    op_un_re : forall {x} -> op x un = x
\end{verbatim}
\normalsize
\begin{block}<2->{The Catch}
  In the signature of \texttt{fold} \dots
  \begin{itemize}
  \item \texttt{E} is \textbf{not} a \texttt{Set}, but an element of a
    data type that encodes Haskell datatypes
  \item the type of the operation is
    \texttt{(Exp E -> Exp E -> Exp E)}\\
    a HOAS encoding of the operation's \textbf{syntax}
  \item<3-> Associativity of syntax is meaningless \dots 
  \end{itemize}

\end{block}
\end{frame}
\end{document}

>>>>

\begin{frame}
  \frametitle{Introduction}
  \begin{itemize}
  \item Domain-specific embedded languages
  \item Dependent types
  \item D-Generative programming
  \item Data parallelism
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Overview}
  \begin{block}{Mission}
    Construct a dependently-typed, generative, data parallel programming language
  \end{block}
  \begin{block}<2->{Means}
    \dots\ by creating an Agda frontend for the Accelerate language.
  \end{block}
\end{frame}

\section{Accelerate}
\begin{frame}
  \frametitle{The Accelerate Language}
  \begin{itemize}
  \item Data-parallel array language
  \item Expressive embedding in Haskell
  \item Generates GPU code at run time
  \item See Chakravarty et al, DAMP2011
  \end{itemize}
\end{frame}

@@

\begin{frame}[fragile]
  \frametitle{Types in Accelerate}
  \begin{itemize}
  \item \texttt{Vector} and \texttt{Scalar} are suitable
    instantiations of a parameterized array type:
\begin{verbatim}
type Scalar e = Array DIM0 e
type Vector e = Array DIM1 e
\end{verbatim}
  \item Dimensions in turn are abbreviations for certain array shapes:
\begin{verbatim}
type DIM0 = Z
type DIM1 = DIM0 :. Int
-- <and so on>
\end{verbatim}
  \item Shapes are built from two basic constructors
\begin{verbatim}
data Z       = Z
data sh :. i = sh :. i
\end{verbatim}
  \end{itemize}
\end{frame}

@@

\begin{frame}[fragile]
  \frametitle{Indexed Types}
  \begin{itemize}
  \item Standard example: vector type indexed by its size
\begin{verbatim}
data Vec (A : Set) : Nat -> Set where
  []   : Vec A zero
  _::_ : {n : Nat} -> A -> Vec A n -> Vec A (suc n)
\end{verbatim}
  \item<2-> Enables the construction of statically safe access operations
    for vectors
\begin{verbatim}
get : forall {A : Set} {n : Nat} ->
      Vec A n -> (m : Nat) -> m < n -> A
get []        _       ()      -- impossible case
get (x :: xs) zero    p       = x
get (x :: xs) (suc m) (s<s p) = get xs m p
\end{verbatim}
  \end{itemize}
\end{frame}

@@

\begin{frame}[fragile]
  \frametitle{Example: Dot Product in Agda-Accelerate}
  \vspace{-\baselineskip}
\begin{verbatim}
dotp : forall {n : Nat}
     -> PreVector n Float -> PreVector n Float
     -> Scalar Float
dotp xs ys = 
  let xs' = use xs
      ys' = use ys
  in
  fold _+_ ("0" ::: Float) (zipWith _*_ xs' ys')
\end{verbatim}
  \begin{itemize}
  \item \texttt{PreVector n Float} : a vector of $n$ elements of
    type \texttt{Float}
  \item \texttt{Scalar Float} : \textbf{code} that evaluates to a scalar of type \texttt{Float}
  \end{itemize}
\end{frame}
\begin{frame}[fragile]
  \frametitle{The Array Type}
  \begin{itemize}
  \item The array type is indexed over a shape
  \item Shapes contain size information
\begin{verbatim}
data Shape : Set where
  Z     : Shape
  _:<_> : Shape -> Nat -> Shape
\end{verbatim}
  \item \texttt{Vector} and \texttt{Scalar} are, again,
    specializations of a shape-indexed array type
\begin{verbatim}
Vector n E    = Array    (Z :< n >) E
Scalar E      = Array    Z          E
\end{verbatim}
  \end{itemize}
\end{frame}
\begin{frame}[fragile]
  \frametitle{Bounds Checking: zipWith}
\begin{verbatim}
zipWith : forall {A B C} {sh}
        -> (Exp A -> Exp B -> Exp C)
        -> Array sh A -> Array sh B -> Array sh C
\end{verbatim}
  \begin{itemize}
  \item Ensures that only arrays of the exact same shape and size are
    zipped together
  \item More restrictive than Accelerate that only checks the
    dimension and then takes the minimum
  \item But that can also be encoded
  \end{itemize}
\end{frame}
\begin{frame}[fragile]
  \frametitle{Accelerate's zipWith Function in Agda}
\begin{verbatim}
zipWith' : forall {A B C} {sha shb} {p : EqShape sha shb}
        -> (Exp A -> Exp B -> Exp C)
        -> Array sha A -> Array shb B
        -> Array (shmin sha shb {p}) C
\end{verbatim}
where\scriptsize
\begin{verbatim}
data EqShape : Shape -> Shape -> Set where
  EqZ : EqShape Z Z
  EqS : {s1 s2 n1 n2} -> EqShape s1 s2 -> EqShape (s1 :< n1 >) (s2 :< n2 >)

@@

\begin{frame}[fragile]
  \frametitle{Bounds Checking: fold and fold1}
  \vspace{-\baselineskip}
  \begin{itemize}
  \item \texttt{fold} and \texttt{fold1} take a binary operation and
    reduce one dimension of an array using the operation
\begin{verbatim}
fold op e []          = e
fold op e [a]         = a
fold op e [a1 ... an] = op (fold op e [a1 ... ai])
                           (fold op e [ai+1 ... an])
\end{verbatim}
  \item \texttt{fold} works for any array, but \texttt{fold1} elides
    the \texttt{e} argument and requires a non-empty array
  \item The latter restriction is not expressed in Haskell-Accelerate
  \end{itemize}
  \begin{block}<2->{\vspace{-\baselineskip}}
\vspace{-\baselineskip}\small
\begin{verbatim}
fold  :: (Shape ix, Elt a) =>
         (Exp a -> Exp a -> Exp a) -> Exp a ->
         Acc (Array (ix :. Int) a) -> Acc (Array ix a)
fold1 :: (Shape ix, Elt a) =>
         (Exp a -> Exp a -> Exp a) ->
         Acc (Array (ix :. Int) a) -> Acc (Array ix a)
\end{verbatim}
\vspace{-\baselineskip}
  \end{block}
\end{frame}

@@

fold1 : forall {E}{sh}{n}
      -> (Exp E -> Exp E -> Exp E)
      -> Array (sh :< suc n >) E
      -> Array sh E
\end{verbatim}
\end{frame}
\begin{frame}[fragile]
  \frametitle{Another Issue with fold and fold1}
  \begin{itemize}
  \item \texttt{fold op e} and \texttt{fold1 op}  partition their array
    argument and apply \texttt{op} in a bracketing that is a-priori unknown
  \item For deterministic results, \texttt{op} must be associative and
    (in case of \texttt{fold}) \texttt{e} should be a unit of \texttt{op}
  \item<2-> Such restrictions can be expressed in Agda
  \item<3-> Might ask for a signature like this
\begin{verbatim}
fold : ... -> Monoid E
     -> Array (sh :< n >) E -> Array sh E
\end{verbatim}
  \end{itemize}
\end{frame}
\begin{frame}[fragile]
  \frametitle{A Monoidal Catch}
An encoding of monoids in Agda:\scriptsize
\begin{verbatim}
record Monoid (E : Set) : Set where
  field
    op : E -> E -> E
    un : E
    op_assoc : forall {x y z} -> op (op x y) z = op x (op y z)
    op_un_le : forall {x} -> op un x = x
    op_un_re : forall {x} -> op x un = x
\end{verbatim}
\normalsize
\begin{block}<2->{The Catch}
  In the signature of \texttt{fold} \dots
  \begin{itemize}
  \item \texttt{E} is \textbf{not} a \texttt{Set}, but an element of a
    data type that encodes Haskell datatypes
  \item the type of the operation is
    \texttt{(Exp E -> Exp E -> Exp E)}\\
    a HOAS encoding of the operation's \textbf{syntax}
  \item<3-> Associativity of syntax is meaningless \dots 
  \end{itemize}
\end{block}
\end{frame}

\section{Limitation}

\begin{frame}[fragile]
  \frametitle{A Limitation}
  \begin{itemize}
  \item Haskell-Accelerate admits implementing a \texttt{filter}
    operation (for vectors) with the usual meaning
  \item<2-> Surprisingly, it cannot easily be implemented in Agda-Accelerate!
  \item<2-> Why is that?
  \end{itemize}
  \begin{block}<3->{Implementation attempt}
\begin{verbatim}
filter : forall {n m : Nat}{E : Element}
       -> Vector n E -> (Exp E -> Exp Bool) 
       -> Vector m E
\end{verbatim}
    \begin{itemize}
    \item Problem: What is \texttt{m}?
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}[fragile]
  \frametitle{A Workaround}
  \begin{itemize}
  \item Provide a special type of filtered vectors that pair each
    element with a boolean indicating its presence
\begin{verbatim}
FVector : Nat -> Element -> Set
FVector n E = Vector n (Pair Bool E)
\end{verbatim}
  \item<2-> Now filtering becomes straightforward 
\begin{verbatim}
filterF : forall {n : Nat}{E : Element}
        -> (Exp E -> Exp Bool)
        -> FVector n E -> FVector n E
filterF {n}{E} pred vec =
  map g vec
  where
  g : Exp (Pair Bool E) -> Exp (Pair Bool E)
  g bx = pair ((fst bx) && p (snd bx)) x
\end{verbatim}
  \end{itemize}
\end{frame}
\begin{frame}[fragile]
  \frametitle{More Operations on Filtered Vectors}
  \vspace{-\baselineskip}
  \begin{block}{Mapping}
  \vspace{-\baselineskip}\small
\begin{verbatim}
mapF : forall {n : Nat}{E F : Element}
     -> Exp F -> (Exp E -> Exp F) 
     -> FVector n E -> FVector n F
\end{verbatim}
    requires a dummy element of \texttt{F} to put in the result when
    the corresponding source entry is not present
  \end{block}
  \begin{block}<2->{Folding}
  \vspace{-\baselineskip}\small
\begin{verbatim}
foldF : forall {n : Nat}{E : Element}
      -> (Exp E -> Exp E -> Exp E) -> Exp E
      -> FVector n E -> Scalar E
foldF f e vec =
  fold f e (map (\ bx -> if (fst bx) then (snd bx) else e) vec)
\end{verbatim}
    gets out of the filtered vector back to a ``real'' value
  \end{block}
\end{frame}

\section{Wrap Up}

\begin{frame}
  \frametitle{In the paper}
  \begin{itemize}
  \item More information on the representation of types and constants
  \item Further examples
  \item Description of the implementation using the foreign function
    interface of the Alonzo compiler
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Conclusion}
  \begin{itemize}
  \item Only partially successful
  \item Bounds checking works well
  \item Large part could be done using lifted datatypes
  \item Generative languages do not fit well with dependent types
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Example: Polymorphic Dot Product in Agda-Accelerate}
  \vspace{-\baselineskip}
\begin{verbatim}
dotp : forall {E : Element} {{p : IsNumeric E}} {n : Nat}
     -> PreVector n E -> PreVector n E -> Scalar E
dotp{E} xs ys = 
  let xs' = use xs
      ys' = use ys
  in
  fold _+_ ("0" ::: E) (zipWith _*_ xs' ys')
\end{verbatim}
  \begin{itemize}
  \item \texttt{PreVector n E} : a vector of $n$ elements of
    type \texttt{E}
  \item \texttt{Scalar E} : code that evaluates to a scalar of type \texttt{E}
  \item \texttt{IsNumeric E} restricts \texttt{E} to numeric types
  \end{itemize}
\end{frame}

\end{document}
